Ecosystem size and complexity dictate riverine biodiversity

### 2021-02-03 manuscript.Rmd
edits: minor updates on mansucript

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the pervasive conflict between biodiversity conservation and societal demands.

bibliography: reference.bib
csl: nature.csl
---

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000a]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesArea2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008a], larger metacommunity size[@ben-hurDisentanglingMechanismsUnderlying2020;@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts. For example, conservation ecologists have used the SAR estimates to design marine and terrestrial protected areas, which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a], should control the ecosystem's habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial turnover of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation exists (i.e., species travel short distances). This pattern reflects significant spatial turnover of species composition over the branching network. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects)[@thompsonProcessbasedMetacommunityFramework2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how the apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching may not contribute to the ecosystem's habitat heterogeneity because local environmental noise masks environmental differences between tributaries. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat heterogeneity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes[@altermattDiversityRiverineMetacommunities2013; @bendaNetworkDynamicsHypothesis2004; @mooreEmergentStabilityLarge2015; @nakamuraDisturbanceRegimesStream2000; @petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. Furthermore, the observed patterns were closely aligned with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, the fraction of forest, and dam density) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial turnover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties[@bendaNetworkDynamicsHypothesis2004]. Although our analysis is correlative and cannot provide conclusive evidence, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turnerLandscapeEcologyTheory2015] and aquatic ecosystems[@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the pervasive conflict between biodiversity conservation and societal demands. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\mu_z, \Omega_z)$. $\mu_z$ is the vector of mean environmental conditions of habitat patches, and $\Omega_z$ is the variance-covariance matrix. Spatial autocorrelation in temporal environmental dynamics is considered by expressing the off-diagonal elements as:

@@

$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path [measured as the number of habitat patches] to reach a target patch). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

We constructed two models for each of the diversity metrics: global and region-specific models. The global model assumes that the effects of watershed area and branching probability are consistent across biogeographic regions (i.e., no interactive effects of the region with other linear predictors).

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesize that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two geographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.
bibliography: reference.bib
csl: nature.csl
---

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesareaRelationship2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008], larger metacommunity size[@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts[@myersBiodiversityHotspotsConservation2000; @desmetUsingSpeciesareaRelationship2004;@neigelSpeciesareaRelationshipsMarine2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas[@desmetUsingSpeciesareaRelationship2004; @neigelSpeciesareaRelationshipsMarine2003], which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turnerLandscapeEcologyTheory2015;@rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@carraraDendriticConnectivityControls2012; @faganConnectivityFragmentationExtinction2002; @teruiMetapopulationStabilityBranching2018;@brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldoRiverNetworksEcological2018;@petersonModellingDendriticEcological2013]. For example, this research has highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carraraDendriticConnectivityControls2012]. However, most research has explored the consequences of branching complexity for local community structure[@brownDendriticNetworkStructure2010; @carraraDendriticConnectivityControls2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], should control the ecosystem's habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial turnover of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries[@mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation exists (i.e., species travel short distances). This pattern reflects significant spatial turnover of species composition over the branching network. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects)[@thompsonProcessbasedMetacommunityFramework2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching may not contribute to the ecosystem's habitat heterogeneity because local environmental noise masks environmental differences between tributaries. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat heterogeneity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, and flow/sediment regimes[@petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two geographic regions. Furthermore, the observed patterns were closely aligned with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, the fraction of forest, and dam density) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Field studies (e.g., mark-recapture) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters[@radingerPatternsPredictorsFish2014]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial turnover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties. Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @grillMappingWorldFreeflowing2019], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turnerLandscapeEcologyTheory2015] and aquatic ecosystems[@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\boldsymbol{\mu_z}, \boldsymbol{\Omega_z})$, where $\boldsymbol{\mu_z}$ is the vector of mean environmental conditions of habitat patches and $\boldsymbol{\Omega_z}$ is the variance-covariance matrix. The diagonal elements of the variance-covariance matrix $\sigma^2_z$ represent the degree of temporal environmental variation. The off-diagonal elements $\Omega_{xy}$ represent spatial autocorrelation in temporal environmental variation as:

@@

$\Omega_{xy}$ denotes the temporal covariance of environmental conditions between patch $x$ and $y$, which is assumed to decay exponentially with increasing distance between the patches $d_{xy}$ (the shortest network path measured as the number of habitat patches). The parameter $\phi$ determines the degree of the distance decay of environmental correlates.

@@

We constructed two models for each of the diversity metrics: global and region-specific models. The global model assumes that the effects of watershed area and branching probability are consistent across geographic regions (i.e., no interactive effects of the region with other linear predictors).
